---
title: "Fine-Grained Energy Demand" 
author: "Avery Rogers"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(lubridate)

# Parameters

  # Day-ahead demand forecast API
balance_data <- ("~/GitHub/EIA930_BALANCE_2020_Jan_Jun.csv")

  # Variable types
vars_types <- 
  cols_only(
    `Balancing Authority` = col_character(),
    `Data Date` = col_character(),
    `Hour Number` = col_character(),
    `Demand Forecast (MW)` = col_character(),
    `Demand (MW)` = col_character(),
    `Net Generation (MW)` = col_character(),
    `Total Interchange (MW)` = col_character(),
    `Net Generation (MW) from Coal` = col_character(),
    `Net Generation (MW) from Natural Gas` = col_character(),
    `Net Generation (MW) from Nuclear` = col_character(),
    `Net Generation (MW) from Solar` = col_character(),
    `Net Generation (MW) from Wind` = col_character()
  )

  # Vector of top balancing authorities
TOP_AUTHORITIES <- c("PJM", "MISO", "SOCO", "ERCO", "SWPP", "CISO")


#===============================================================================

# Code
```


# Energy Demand and Generation at Top Electricity Balancing Authorities

For this portion of my C01, I wanted to explore the different electricity balancing authorities across the United States and understand how they generate enrgy on a more granular level, as well as how they calibrate their forecasts with demand and generation. Here, I will focus on the 6 largest balancing authorities, which operate most of the US electricity markets: 

* PJM: Standing for Pennsylvania-New Jersey-Maryland Connection, the PJM serves the Mid-Atlantic region and some of the Northeast.
*MISO: Stanfind for Midcontinent Independent System Operator, MISO serves the upper Midwest and parts of the South. 
* SOCO: Standing for Southern Company Services, SOCO serves the majority of the American South. 
* ERCO: Standing for Electric Reliability Council of Texas, ERCO serves the state of Texas.
* SWPP: Standing for Southwest Power Pool, SWPP serves the American Southwest. 
* CISO: Standing for California Independent Service Operator, CISO serves the state of California. 

## Tidying the Data

```{r}
balance <- 
  balance_data %>% 
  read_csv(col_types = vars_types) %>% 
  transmute(
    balancing_authority = (`Balancing Authority`),
    datetime = mdy_h(paste(`Data Date`, `Hour Number`)),
    demand_forecast = `Demand Forecast (MW)`,
    demand = `Demand (MW)`,
    net_generation = `Net Generation (MW)`,
    total_interchange = `Total Interchange (MW)`,
    net_gen_coal = `Net Generation (MW) from Coal`,
    net_gen_natgas = `Net Generation (MW) from Natural Gas`,
    net_gen_nuclear = `Net Generation (MW) from Nuclear`,
    net_gen_solar = `Net Generation (MW) from Solar`,
    net_gen_wind = `Net Generation (MW) from Wind`
  ) %>% 
  filter(balancing_authority %in% TOP_AUTHORITIES) %>% 
  mutate_at(vars(demand_forecast:net_gen_wind), ~ as.integer(str_remove(., ",")))
  
balance
```

## Net Generation by Balancing Authority

Here, I compare the net generation on January 1st and 2nd, 2020 for each operator. 

```{r}
balance  %>% 
  filter(datetime < "2020-01-03 0:00:00") %>% 
  ggplot(aes(datetime, net_generation, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "6 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts)",
    title = "Electricity generation by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
  
```

As we can see, PJM is operates the most electricity of all the authorities, followed by MISO and the rest with significantly less electricity under their control. All of the authorities see a jump in generation at about 4PM, as would be expected given that the sun begins to set at about this time in the winter. It then tapers off as people go to sleep and spikes again around 5AM, when people begin to rise and start their days. 

## Balancing Authority by Energy Source

Here, I look at each source of energy tracked by the Energy Information Administration - natural gas, coal, nuclear, solar, and wind, and their usage levels across the balance authorities.

### Coal

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_coal, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Coal",
    title = "Electricity generation from coal by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Nuclear

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_nuclear, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Nuclear",
    title = "Electricity generation from nuclear by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Natural Gas

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_natgas, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Natural Gas",
    title = "Electricity generation from natural gas by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Solar

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_solar, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Solar",
    title = "Electricity generation from solar by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```



### Wind

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_wind, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Wind",
    title = "Electricity generation from wind by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

## Renewable Energy Reliability Around the Clock

For this section, I want to look at wind, solar, and nuclear at the balancing authorities where they are most common to see how they do as far as reliability over the course of many days. The patterns that emerge will be instructive for understanding the possible energy mixes of a greener future with reliable electricity. 

### Solar

```{r}
balance  %>% 
  filter(balancing_authority == "CISO") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_solar)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_solar), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Solar",
    title = "Hourly electricity generation from solar by CISO"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

As we can see, solar has similar peaks in its electricity productivity each day in California, where sunshine is fairly abundant and regular. However, as is to be expected, solar capacity drops to zero in the late hours of the day until the sun is back up. To get a more complete sense of this picture, let's zoom in on a day: 

```{r}
balance  %>% 
  filter(balancing_authority == "CISO") %>% 
  filter(date(datetime) == "2020-01-01") %>% 
  ggplot(aes(datetime, net_gen_solar)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_solar), 
    color = "darkblue",
    size = 4
  ) +
  scale_x_datetime(breaks = "3 hours") +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Time on January 1st",
    y = "Net Generation (Megawatts) From Solar",
    title = "Hourly electricity generation from solar by CISO on Jan 1st, 2020"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

As we see here, solar only generates any electricity from 8AM - 4PM in the winter in California, with the builk between 9AM and 3PM. 

### Wind

```{r}
balance  %>% 
  filter(balancing_authority == "ERCO") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_wind)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_wind), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Wind",
    title = "Hourly electricity generation from wind by ERCO"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

Looking at wind electricity productivity in Texas under ERCO, we see that its productivity does not have the same daily peaks and falls as solar energy does in California, but rather has an unpredictable pattern across days depending on weather conditions. 

### Nuclear

```{r}
balance  %>% 
  filter(balancing_authority == "PJM") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_nuclear)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_nuclear), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Nuclear",
    title = "Hourly electricity generation from nuclear by PJM"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

Nuclear, as we see from PJM, is by far the most stable source of renewable energy documented by the EIA across balance authorities. It has nearly perfect consistency across all hours of the day, every day of the week. 

