---
title: Energy Consumption Sources by State
author: Avery Rogers
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(ggalluvial)
library(lubridate)

# Parameters

  # Energy consumption data from EIA
energy_data <- "~/Downloads/use_all_btu.csv"

  # State population data from FRED
population_data <- "~/Downloads/state_population.csv"

# Day-ahead demand forecast API
balance_data <- ("~/GitHub/EIA930_BALANCE_2020_Jan_Jun.csv")

# MSN codes to keep
msn_to_keep <- 
  c(
    "CLTCB", 
    "GETCB", 
    "HYTCB", 
    "NGTCB", 
    "NUETB", 
    "SOTCB", 
    "WWTCB",
    "WYTCB"
  )

# Recode MSN values
recode_msn <- 
  c(
    "CLTCB" = "Coal",
    "GETCB" = "Geothermal",
    "HYTCB" =  "Hydropower",
    "NGTCB" = "Natural Gas", 
    "NUETB" = "Nuclear", 
    "SOTCB" = "Solar",
    "WWTCB" = "Wood and Waste", 
    "WYTCB" = "Wind"
  ) 

  # Variable types for balancing authority data
vars_types <- 
  cols_only(
    `Balancing Authority` = col_character(),
    `Data Date` = col_character(),
    `Hour Number` = col_character(),
    `Demand Forecast (MW)` = col_character(),
    `Demand (MW)` = col_character(),
    `Net Generation (MW)` = col_character(),
    `Total Interchange (MW)` = col_character(),
    `Net Generation (MW) from Coal` = col_character(),
    `Net Generation (MW) from Natural Gas` = col_character(),
    `Net Generation (MW) from Nuclear` = col_character(),
    `Net Generation (MW) from Solar` = col_character(),
    `Net Generation (MW) from Wind` = col_character()
  )

# Vector of top balancing authorities
TOP_AUTHORITIES <- c("PJM", "MISO", "SOCO", "ERCO", "SWPP", "CISO")

#===============================================================================


```

# Part I: Energy Consumption for Electricity by State and Year

In this section, I am only looking at the sources of energy that primarily contribute to electricity generation across states: coal, geothermal power, hydropower, natural gas, nuclear power, solar power, wood and waste power, and wind power. I remove all the other energy usage statistics from the dataset to hone in on these. Most notably, I remove petroleum, which is the largest source of energy in many states but primarily for transportation rather than electriticy in the grid. 

An important note about this data: all energy usage is measured in BTU, or British Thermal Units. One BTU is the amount of energy needed to raise one pound of water by one degree Fahrenheit. (I'm not sure why it's called _British_ thermal units if the measurements it uses are imperial!). One BTU is approximately 1,055 joules. BTU is the unit used to sell natural gas in the US (dollars per million BTUs), and is the best unit of measurement for comparing different energy sources. 

## Cleaning Data


```{r}
energy_data_clean <- 
  read_csv(energy_data) %>% 
  rename_all(str_to_lower) %>% 
  select(-data_status) %>% 
  pivot_longer(c("1960":"2017"), names_to = "year", values_to = "amount") %>% 
  mutate(year = as.integer(year)) %>% 
  select(state, year, msn, amount) %>% 
  filter(msn %in% msn_to_keep) %>% 
  mutate_at(vars(msn), recode, !!! recode_msn) %>% 
  mutate(amount = replace_na(amount, 0))

energy_data_clean
```

## Getting percentages by source for each year and state

```{r}
consumption_fractions <- 
  energy_data_clean %>% 
  group_by(state, year) %>% 
  mutate(total_cons = sum(amount)) %>% 
  ungroup() %>% 
  mutate(frac_amount = amount / total_cons) %>% 
  select(state, year, msn, frac_amount)

consumption_fractions
```


## Creating necessary plotting functions

```{r}
plot_fractions <- function(state_abbv, state) {
  consumption_fractions %>% 
    filter(state == state_abbv) %>% 
    ggplot(aes(x = year, y = frac_amount, alluvium = msn)) + 
    geom_alluvium(
      aes(fill = msn, color = msn), 
      width = 1/16, 
      alpha = 1/2, 
      decreasing = NA
    ) + 
    scale_x_continuous(
      breaks = seq(1960, 2015, 5),
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      expand = c(0, 0),
      breaks = seq(0, 1, 0.1),
      minor_breaks = NULL,
      labels = scales::percent_format(accuracy = 1)
    ) +
    labs(
      x = "Year",
      y = "Percentage of Energy Consumption",
      title = str_glue("Energy Consumption by Source in {state}, Percentage"),
      caption = "Source: US Energy Information Administration",
      color = "Energy\nSource",
      fill = "Energy\nSource"
    ) +
    theme(
      panel.background = element_rect(fill = "grey95"),
      panel.grid.major = element_line(color = "white"),
      panel.grid.minor = element_line(color = "white")
    )
} 

plot_amount <- function(state_abbv, state) {
  energy_data_clean %>% 
    filter(state == state_abbv) %>% 
    ggplot(aes(x = year, y = amount, alluvium = msn)) + 
    geom_alluvium(
      aes(fill = msn, color = msn), 
      width = 1/16, 
      alpha = 1/2, 
      decreasing = NA
    ) + 
    scale_x_continuous(
      breaks = seq(1960, 2015, 5),
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      expand = c(0, 0),
      labels = scales::number_format(accuracy = 1)
    ) +
    labs(
      x = "Year",
      y = "Amount (billion BTU)" ,
      title = str_glue("Energy Consumption by Source in {state}, Total Amount"),
      caption = "Source: US Energy Information Administration",
      color = "Energy\nSource",
      fill = "Energy\nSource"
    ) +
    theme(
      panel.background = element_rect(fill = "white"),
      panel.grid.major = element_line(color = "gray92"),
      panel.grid.minor = element_line(color = "gray92")
    )
}
```

## Example plot: Alaska

The Alaska plots demonstrate why it is interesting and important to plot not only the percentages of electricity-generating energy sources by year, but also the total amount of electricity-generating enegy sources by year. As we can see in the first plot, Alaska repidly transitions from minimal to very high usage of natural gas in the 1960s and 70s as a percentage of its energy sources. Looking at the second plot, we see a more nuanced story: it was not that Alaska used large amount of coal in the 1960s, but rather that their energy usage overall more than doubled from 1960 to 1970, and continued to grow rapidly in the decades afterwards. 

```{r}
plot_fractions("AK", "Alaska")
plot_amount("AK", "Alaska")
```

## Greatest Consumers By Fraction

```{r}
greatest_consumer <- function(target_msn, target_year) {
  consumption_fractions %>% 
    filter(msn == target_msn, year == target_year) %>% 
    mutate(state = fct_reorder(state, frac_amount)) %>%
    ggplot(aes(state, frac_amount)) + 
    geom_point() +
    coord_flip() +
    scale_y_continuous(
      labels = scales::percent_format(accuracy = 1)
    ) +
    labs(
      x = "State",
      y = str_glue("Percentage of Energy Consumption from {target_msn}"), 
      title = str_glue(
        "Comparative Energy Consumption from {target_msn} in {target_year}",
      caption = "Source: US Energy Information Administration"
      )
    ) +
    theme(
      panel.background = element_rect(fill = "grey98"),
      panel.grid.major = element_line(color = "grey92"),
      panel.grid.minor = element_line(color = "grey92")
    )
} 
```

## Example: Wood and Waste 2017

The northeastern states of Maine, New Hampshire, and Vermont consume a large fraction of energy sourced from wood and waste as of 2017. Who knew!

```{r, fig.height = 7}
greatest_consumer("Wood and Waste", 2017)
```


# Part II: Energy Usage by Population

In this section, I wanted to understand the relationship between energy usage and state population and see whether increases in energy usage were accelerating faster or slower than overall population growth by state. This will help shed more light on how energy consumption is changing per capita and whether or not we are headed in the right direction in terms of sustainability. 

## Cleaning Population Data

```{r}
population_data_clean <- 
  read_csv(population_data) %>% 
  pivot_longer(
    cols = c(AKPOP:WYPOP), 
    names_to = "state", 
    values_to = "population"
  ) %>% 
  mutate_at(vars(state), ~ str_sub(., 1, 2)) %>% 
  mutate(
    year = as.integer(year(DATE)),
    population_thousands = as.integer(population)
  ) %>% 
  select(year, state, population_thousands) %>% 
  filter(year > 1959, year < 2018)

population_data_clean
```

## Combining The Tibbles

Here, I add a variable `amount_per_capita` to see how much energy is being used per each 1,000 people in the state. 

```{r}
total_energy_population <- 
  energy_data_clean %>% 
  left_join(population_data_clean, by = c("year", "state")) %>% 
  mutate(amount_per_capita = amount / population_thousands)

fraction_energy_population <- 
  consumption_fractions %>% 
  left_join(population_data_clean, by = c("year", "state"))
```


## Total Energy Consumption Per Capita by State over Time

When comparatively plotting the total energy consumption per 1,000 people by state over time, we see very surprising results: some states are using as much as 3x the amount of energy per capita as others. 

```{r}
per_capita_energy <- function(state_abbv) { 
  total_energy_population %>% 
    filter(state %in% state_abbv) %>% 
    count(state, year, wt = amount_per_capita) %>% 
    ggplot(aes(year, n, color = state)) +
    geom_line() +
    scale_x_continuous(
      breaks = seq(1960, 2020, 5), 
      expand = c(0, 0)
    ) +
    labs(
      x = "Year", 
      y = "Total Energy Consumption in BTUs per 1,000 People",
      title = "Comparing total energy use per capita in US states",
      color = "State",
      caption = "Sources: St. Louis Fred, US Energy Information Administration"
    ) +
    theme(
      panel.background = element_rect(fill = "white"),
      panel.grid.major = element_line(color = "grey92"),
      panel.grid.minor = element_line(color = "grey92")
    )
}
```

## Example States

As we can see, variation among states' per-capita energy consumption is high, with a state like Kentucky consuming nearly three times as many BTUs of energy per 1,000 people as California. 

```{r}
per_capita_energy(c("FL", "CA", "CT", "CO", "PA", "KY"))
```


## State energy consumption per capita by source 

Here, I look at at comparing the use of different energy sources per 1,000 people within a state. It's particularly interesting to see how sources have substituted over time, and whether or not the overall trend is increasing or decreasing. This supplements the total usage by source graphs above by paying closer attention to the usage standardized for population changes. 

```{r}
per_capita_compare <- function(state_abbv, state, msn_vars) {
  total_energy_population %>% 
  filter(state == state_abbv, msn %in% msn_vars) %>% 
  ggplot(aes(year, amount_per_capita, color = msn)) +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1960, 2020, 5), 
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = c(0, 0)) + 
  labs(
    x = "Year", 
    y = "Total Energy Consumption per 1,000 People (BTUs)",
    title = str_glue("{state} energy use per capita by energy source"),
    color = "Energy\nSource",
    caption = "Sources: St. Louis FRED, US Energy Information Administration"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
}
```

## Example: Maine

We see a sudden spike in natural gas consumption per capita in Maine that trails off after its peak. Hydropower and wood and waste power stay relatively constant after 1980. 

```{r}
per_capita_compare(
  "ME", 
  "Maine", 
  c("Coal", "Natural Gas", "Wood and Waste", "Hydropower")
)
```




# Part III: Energy Demand and Generation at Top Electricity Balancing Authorities

For this portion of my C01, I wanted to explore the different electricity balancing authorities across the United States and understand how they generate enrgy on a more granular level, as well as how they calibrate their forecasts with demand and generation. Here, I will focus on the 6 largest balancing authorities, which operate most of the US electricity markets: 

* PJM: Standing for Pennsylvania-New Jersey-Maryland Connection, the PJM serves the Mid-Atlantic region and some of the Northeast.
*MISO: Stanfind for Midcontinent Independent System Operator, MISO serves the upper Midwest and parts of the South. 
* SOCO: Standing for Southern Company Services, SOCO serves the majority of the American South. 
* ERCO: Standing for Electric Reliability Council of Texas, ERCO serves the state of Texas.
* SWPP: Standing for Southwest Power Pool, SWPP serves the American Southwest. 
* CISO: Standing for California Independent Service Operator, CISO serves the state of California. 

## Tidying the Data

```{r}
balance <- 
  balance_data %>% 
  read_csv(col_types = vars_types) %>% 
  transmute(
    balancing_authority = (`Balancing Authority`),
    datetime = mdy_h(paste(`Data Date`, `Hour Number`)),
    demand_forecast = `Demand Forecast (MW)`,
    demand = `Demand (MW)`,
    net_generation = `Net Generation (MW)`,
    total_interchange = `Total Interchange (MW)`,
    net_gen_coal = `Net Generation (MW) from Coal`,
    net_gen_natgas = `Net Generation (MW) from Natural Gas`,
    net_gen_nuclear = `Net Generation (MW) from Nuclear`,
    net_gen_solar = `Net Generation (MW) from Solar`,
    net_gen_wind = `Net Generation (MW) from Wind`
  ) %>% 
  filter(balancing_authority %in% TOP_AUTHORITIES) %>% 
  mutate_at(vars(demand_forecast:net_gen_wind), ~ as.integer(str_remove(., ",")))
  
balance
```


## Checking if Demand, Generation, and Interchange Data Add Up

```{r}
balance %>% 
  filter(demand != net_generation - total_interchange)

balance %>% 
  mutate(difference = demand - net_generation + total_interchange) %>% 
  select(difference) %>%
  summary
  
```

We can see here that 4,173 of the 6,912 entries in our data do not have net generation minus total interchange equal to demand. In other words, most of our entries do not have balanced energy production and energy consumption. Looking more closely at the distribution of these values, we see that the median difference between generation and consumption is 0, though the mean is 327 MW. Stunningly, the minimum difference is -8,629 MW, and the maximum is 13,612 MW. This isn't great news for our analysis. Let's see if there are any particular facilities that aren't adding up:

```{r}
balance %>% 
  filter(demand != net_generation - total_interchange) %>% 
  count(balancing_authority)
```

It looks like CISO, MISO, and PJM have the most errors, with ERCO and SWPP following. SOCO has almost no errors in its reporting. Unfortunately, the EIA website does not offer much guidance on these errors. They provide a disclaimer at the bottom of their balancing authority data that "The information submitted by reporting entities is preliminary data and is made available "as-is" by EIA. Neither EIA nor reporting entities are responsible for reliance on the data for any specific use."


## Net Generation by Balancing Authority

Here, I compare the net generation in January, 2020 for each operator. 

```{r}
balance  %>% 
  filter(month(datetime) == 1) %>% 
  group_by(balancing_authority) %>% 
  summarize(
    total_gen_jan = sum(net_generation, na.rm = TRUE),
    total_demand_jan = sum(demand, na.rm = TRUE)
  ) %>% 
  pivot_longer(
    cols = c(total_gen_jan, total_demand_jan), 
    names_to = "type_use", 
    values_to = "megawatts"
  ) %>% 
  ggplot(aes(fct_rev(fct_reorder(balancing_authority, megawatts)), megawatts, fill = type_use)) + 
  geom_col(position = "dodge") +
  scale_fill_manual(
    labels = c("Demand", "Net Generation"), 
    values = c("brown1", "skyblue")
  ) +
  labs(
    x = "Balancing Authority",
    y = "Total Generation in January (MW)",
    title = "Generation Levels by Balancing Authority",
    fill = "Energy Information\nType"
  )
  
```

As we can see, PJM is operates the most electricity of all the authorities, followed by MISO and the rest with significantly less electricity under their control. Interestingly, PJM is the only authority that produced more energy in Jan 2020 than it used, ERCO, SWPP, and SOCO were approximately balanced in generation and demand, and MISO and CISO's residents demanded more than the authorities generated in this period. 


## Balancing Authority by Energy Source

Here, I look at each source of energy tracked by the Energy Information Administration - natural gas, coal, nuclear, solar, and wind, and their usage levels across the balance authorities.

### Coal

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_coal, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Coal",
    title = "Electricity generation from coal by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Nuclear

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_nuclear, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Nuclear",
    title = "Electricity generation from nuclear by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Natural Gas

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_natgas, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Natural Gas",
    title = "Electricity generation from natural gas by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

### Solar

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_solar, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Solar",
    title = "Electricity generation from solar by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```



### Wind

```{r}
balance  %>% 
  filter(datetime < "2020-01-02 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_wind, color = balancing_authority)) + 
  geom_line() +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%b %e, %l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Date and Time",
    y = "Net Generation (Megawatts) From Wind",
    title = "Electricity generation from wind by balancing authority",
    color = "Balancing\nAuthority"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

## Renewable Energy Reliability Around the Clock

For this section, I want to look at wind, solar, and nuclear at the balancing authorities where they are most common to see how they do as far as reliability over the course of many days. The patterns that emerge will be instructive for understanding the possible energy mixes of a greener future with reliable electricity. 

### Solar

```{r}
balance  %>% 
  filter(balancing_authority == "CISO") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_solar)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_solar), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Solar",
    title = "Hourly electricity generation from solar by CISO"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

As we can see, solar has similar peaks in its electricity productivity each day in California, where sunshine is fairly abundant and regular. However, as is to be expected, solar capacity drops to zero in the late hours of the day until the sun is back up. To get a more complete sense of this picture, let's zoom in on a day: 

```{r}
balance  %>% 
  filter(balancing_authority == "CISO") %>% 
  filter(date(datetime) == "2020-01-01") %>% 
  ggplot(aes(datetime, net_gen_solar)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_solar), 
    color = "darkblue",
    size = 4
  ) +
  scale_x_datetime(breaks = "3 hours") +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%l%p") +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  labs(
    x = "Time on January 1st",
    y = "Net Generation (Megawatts) From Solar",
    title = "Hourly electricity generation from solar by CISO on Jan 1st, 2020"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

As we see here, solar only generates any electricity from 8AM - 4PM in the winter in California, with the builk between 9AM and 3PM. 

### Wind

```{r}
balance  %>% 
  filter(balancing_authority == "ERCO") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_wind)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_wind), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Wind",
    title = "Hourly electricity generation from wind by ERCO"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

Looking at wind electricity productivity in Texas under ERCO, we see that its productivity does not have the same daily peaks and falls as solar energy does in California, but rather has an unpredictable pattern across days depending on weather conditions. 

### Nuclear

```{r}
balance  %>% 
  filter(balancing_authority == "PJM") %>% 
  filter(datetime < "2020-01-11 0:00:00") %>% 
  ggplot(aes(datetime, net_gen_nuclear)) + 
  geom_segment(
    aes(x = datetime, y = 0, xend = datetime, yend = net_gen_nuclear), 
    color = "darkblue"
  ) +
  scale_x_datetime(date_labels = "%b %e") +
  labs(
    x = "Date",
    y = "Net Generation (Megawatts) From Nuclear",
    title = "Hourly electricity generation from nuclear by PJM"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey92")
  )
```

Nuclear, as we see from PJM, is by far the most stable source of renewable energy documented by the EIA across balance authorities. It has nearly perfect consistency across all hours of the day, every day of the week. 
